---
title: "Data Preprocessing"
author: "Tyler Katz"
date: "2025-07-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(janitor)
```

## LOADING & MERGING DATA

This section loads the primary datasets required for analysis:

  -`train.csv` contains daily sales records for individual Rossmann stores, including features such as sales, customer traffic, promotions, and holidays.

  -`store.csv` provides store-level metadata, including store type, assortment level, and competition indicators.

The two datasets are merged using the `Store` identifier to create a unified dataset (`merged_data`) that combines both transactional and contextual information for each store. This merged dataset forms the foundation for all subsequent preprocessing, feature engineering, and modeling.
```{r}

#Load training and store metadata
train <- read_csv("../Data/train.csv", show_col_types = FALSE)
store <- read_csv("../Data/store.csv", show_col_types = FALSE)

#Preview data
glimpse(train)
glimpse(store)

#Merge training data with store meta data
merged_data <- left_join(train, store, by = "Store")

#Preview merged data
glimpse(merged_data)

```

## HANDLING MISSING VALUES

This section identifies and addresses missing values in the merged dataset to ensure it is ready for modeling. A column-wise summary of missing values is generated before and after processing to validate that all gaps have been addressed.

The following imputation strategies are applied:

  -`CompetitionDistance`: Imputed with the median value, as it is a continuous variable with predictive relevance and no valid default.

  -`CompetitionOpenSinceMonth` and `CompetitionOpenSinceYear`: Imputed with their respective medians to preserve temporal continuity. These values are often missing when competition has not yet opened.

  -`Promo2SinceWeek` and `Promo2SinceYear`: Imputed with -1 to indicate that the store does not participate in the `Promo2` program. These values are only relevant for stores with `Promo2` = 1.

  -`PromoInterval`: Replaced with "None" to explicitly denote the absence of a recurring promotion schedule.

These choices ensure that missingness is handled in a way that maintains both the integrity and interpretability of the data.

```{r}

print("Columns with missing values")
#Check for missing values in each column
sapply(merged_data, function(x) sum(is.na(x)))

#Impute or handle missing values
merged_data <- merged_data %>% mutate(
  
  #Impute competition distance
  CompetitionDistance = ifelse(
    is.na(CompetitionDistance),
    median(CompetitionDistance, na.rm = TRUE),
    CompetitionDistance
    ),
  
  #Impute competition open since month
  CompetitionOpenSinceMonth = ifelse(
      is.na(CompetitionOpenSinceMonth),
      median(CompetitionOpenSinceMonth, na.rm = TRUE),
      CompetitionOpenSinceMonth
    ),
  
  #Impute competition open since year 
  CompetitionOpenSinceYear = ifelse(
      is.na(CompetitionOpenSinceYear),
      median(CompetitionOpenSinceYear, na.rm = TRUE),
      CompetitionOpenSinceYear
    ),
  
  #Add dummy values to indicate no Promo2
  Promo2SinceWeek = ifelse(is.na(Promo2SinceWeek), -1, Promo2SinceWeek),
  Promo2SinceYear = ifelse(is.na(Promo2SinceYear), -1, Promo2SinceYear),
  
  #Add a none value to indiviate no promo intervals
  PromoInterval = replace_na(PromoInterval, "None")
)

print("Columns with missing values after processing")
#Verifies all missing values are handled
sapply(merged_data, function(x) sum(is.na(x)))
```

##FEATURE ENGINEERING

This section extracts and derives new features from the existing Date and Promo columns to support both data visualization and predictive modeling:

  -The `Date` column is converted to a proper date type to enable time-based operations.

  -Calendar components such as `Year`, `Month`, `Day`, and `Week` are extracted to capture seasonal and temporal patterns in sales behavior.

  -The `DayOfWeek` variable is relabeled using descriptive weekday names (Mon–Sun) to enhance interpretability in plots and summaries.

  -A binary `IsWeekend` indicator is created to flag Saturdays and Sundays, which often exhibit different sales dynamics compared to weekdays.

  -The `IsPromo` column transforms the binary Promo variable into a readable categorical label, allowing for clearer comparisons in visualizations and summaries.

These engineered features provide a more intuitive and structured representation of time-based patterns and promotional effects, enhancing both exploratory analysis and model performance.
```{r}

#Creates features for data visualization and modeling
merged_data <- merged_data %>% mutate(
  
  #Converting date column to date type variable
  Date = ymd(Date),
  
  #Extracting elements of date
  Year = year(Date),
  Month = month(Date),
  Day = day(Date),
  Week = week(Date),
  
  #Labeling DayOfWeek column for more effective visualization
  DayOfWeekLabel = factor(
    DayOfWeek,
    level = 1:7,
    labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
  ),
  
  #Creating a weekend indication variable based on day of week labels
  IsWeekend = ifelse(DayOfWeekLabel %in% c("Sat", "Sun"), 1, 0),
  
  #Creating a promotion indicator column based on the binary variable for more effective visualizations
  IsPromo = ifelse(Promo == 1, "Promo", "No Promo")
)
```

##SAVING CLEANED DATA

The fully preprocessed dataset is saved as a CSV file to the `Data/` directory. This file serves as the finalized input for downstream analysis, modeling, or dashboard applications.
```{r}

#Saves clean data to CSV
write_csv(merged_data, "../Data/preprocessed_data.csv")
```
